Fail Faster

### 2025-03-21 paper/paper.tex
fix words

We lay a path forward.
First, we show how to use \textit{multi-stage programming}---a classic technique for 
building fast DSLs “without regret”~\cn{}---to eliminate the overhead of 
many common generator abstractions.
Second, we characterize the impact of \rand choice through a natural experiment 
that isolates the performance contribution of the \rand within a generator, underscoring its role as a first-class design consideration.
Taken together, these insights can be applied to create tomorrow's generator libraries: those that 
combine high-level design with low-level performance.

@@

Of course, simply arguing that the randomness library is on the hot
path for PBT does not guarantee that a faster sampling leads to measurably faster
generation; for that, we need an experiment. The most obvious experiment is to
simply swap out the randomness library of either \bq{} or ScalaCheck with a totally different, faster one. But, as
discussed previously, generators with different generation orders can be
difficult to compare. To get around this, we exploit a ``natural experiment:
the \rand{} that \bq{} uses --- OCaml's \texttt{Splittable\_random} library --- is slow in a way that
can be improved \emph{without} changing its behavior.
In particular,
due to implementation details related to the OCaml garbage collector,
values of the OCaml type \texttt{int64} are not machine words, but rather
\emph{pointers} to machine words. This means
that \emph{all} \texttt{int64} operations (both arithmetic and bitwise) must
allocate memory cells to contain their output, which has a significant performance benefit.
By building a version that uses much faster ``unboxed'' 64-bit integer arithmetic, we can demonstrate just how much faster bugs can be found just by using a more performant randomness library.

@@

\jwc{Unclear what we should call this section}
As we discussed in Section~\ref{section:motiv}, choosing an inefficient randomness library is
another bottleneck for finding bugs fast. While generator libraries by
and large use sensible sources of randomness, they are not explicitly chosen
with performance in mind. Indeed, much faster randomness libraries\cn and fast sources
of entropy \cn exist. \jwc{... more here?}
To demonstrate that faster random sampling can significanty impact bugfinding power, we
use the natural experiment provided by OCaml's inefficient implementation of
SplitMix. By replacing this slow randomness library with a faster but semantically equivalent
implemetnation, we are able to precisely quantify the bugfinding speedup that using a randomness library gives across a range of PBT scenarios.

>>>>

We lay a path forward.
First, we show how to use \textit{multi-stage programming}---a classic technique for 
building fast DSLs “without regret”~\cn{}---to eliminate the overhead of 
many common generator abstractions.
Second, we characterize the impact of \rand choice through a controlled experiment 
that isolates the performance contribution of the \rand within a generator, underscoring its role as a first-class design consideration.
Taken together, these insights can be applied to create tomorrow's generator libraries: those that 
combine high-level design with low-level performance.

@@

Of course, simply arguing that the randomness library is on the hot
path for PBT does not guarantee that a faster sampling leads to measurably faster
generation; for that, we need an experiment. The most obvious experiment is to
simply swap out the randomness library of either \bq{} or ScalaCheck with a totally different, faster one.
But, as discussed previously, \camlname{} and \scalaname{} are designed to be 100\% semantically equivalent replacements for
their unstaged counterparts, ensuring that any bug-finding speedups are solely attributable to generator performance and not lucky seeds.
Choosing a different \rand{} would break this property, making it much more challenging to assess the performance compared to the baseline.
But, as discussed previously, generators with different generation orders can be
difficult to compare \bcp{This was not discussed previously!}. To get around this, we exploit a coincidence:
the \rand{} that \bq{} uses --- OCaml's \texttt{Splittable\_random} library --- is slow in a way that
can be improved \emph{without} changing its behavior.
In particular,
due to implementation details related to the OCaml garbage collector,
values of the OCaml type \texttt{int64} are not machine words, but rather
\emph{pointers} to machine words. This means
that \emph{all} \texttt{int64} operations (both arithmetic and bitwise) must
allocate memory cells to contain their output, which has a significant performance benefit.
By building a version that uses much faster ``unboxed'' 64-bit integer arithmetic, we can demonstrate just how much faster bugs can be found just by using a more performant randomness library.

@@

\jwc{Unclear what we should call this section}
As we discussed in Section~\ref{section:motiv}, choosing an inefficient randomness library is
another bottleneck for finding bugs fast. While generator libraries by
and large use sensible sources of randomness, they are not explicitly chosen
with performance in mind. Indeed, much faster randomness libraries\cn and fast sources
of entropy \cn exist. \jwc{... more here?}
To demonstrate that faster random sampling can significanty impact bugfinding power, we
use the controlled experiment provided by OCaml's inefficient implementation of
SplitMix. By replacing this slow randomness library with a faster but semantically equivalent
implemetnation, we are able to precisely quantify the bugfinding speedup that using a randomness library gives across a range of PBT scenarios.
